Computing likelihoods under the SMC for a general class of ARGs.

### 2023-12-11 paper.tex
update introduction

\subsection{ARGs are awesome, or general intro about approximations}
% approximation: level of detail

@@

Although ARG, ...,
something about the level of detail.

@@

Instead, approximations of the coalescence with recombination (CwR) have  
driven our ability to perform statistical genomic inference. In particular, 
the ability to formulate the inference problem as a hidden Markov model (HMM) 
has proven to be instrumental [references]. 
Two crucial approximations: 2 characteristics: left to right, and discetization 
of state space by emumerating tree topologies and/or discretising time,
then inference can be performed efficiently using well-known HMM algorithms.
To improve the ability to simulate the CwR, McVean and Cardin 
provided an approximation of the CwR that is both Markovian backwards in time 
as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
By sacrificing precision on the our ability to capture long range linkage 
information, the sequentially Markovian coalescent (SMC) 
became the engine of many powerful inference methods [references]. 
Later improvements, SMC', closed the gap between the SMC and the CwR even 
further.
A second key approximation is not a generative model, instead the  
copying model \citep{li_2003_modeling} was specifically designed to 
generate CwR-like patterns. This heuristic approach has been successfully 
applied for demographic inference [and other things: ADD] 
\citep{sheehan_2013_estimating, steinrucken_2018}.

Both approaches have further formed the basis of the recent breakthroughs in 
scalable ARG inference methods [references].

ARGweaver uses 
sample an ARG from the posterior distribution given the observed variational 
data. Although require limiting the set of possible node times to be compatible 
with HMM, ARGs they produce are fully detailed in the sense that recombinations 
are explitly inferred. 
All other methods whether partially relying on
using SMC [ARGneedle] or Li and Stephens [tsinfer, relate] or not [Espallier, ADD] 
are heuristic in nature and % keep models/method separate from level of detail!!

@@

Although all different, all these methods share the fact that the ARGs they infer 
are approximate structures.
% what is an approximate structure: implicit recombination, polytomies or not, ...
They lack the full detail: implicit recombination, ...
In particular, recombination is dealt with implicitly. Recombination events are 
not directly inferred and thus not dated. Tree topology changes from one local 
tree to the next without explicit inference of the one or more recombination 
events required to explain the inferred tree transition.

@@

These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).

>>>>

\subsection{ARGs are awesome, or general intro about approximations}
% approximation: level of detail
* [Importance of ARGs: ARGs as object of theoretical reasoning vs what can be known]\\
* [difference between precision/detail and accuracy. High precision can 
give a false sense of certainty in the provided estimate].

@@

Although ARG, ...,
something about the level of detail.
% easy by which the CwR can be used to simulate is not matched by the usability of the model for inference

@@

Instead, approximations of the coalescence with recombination (CwR) have  
driven our ability to perform statistical genomic inference. Two approximations 
in particular have proven to be instrumental [...]. Tightly linked to their 
ability to reformulate these inference problems as hidden markov Models (HMM) 
and use the well-known efficient HMM-algorithms. 
Both: 1/ Left to right, and 2/ discetization 
of state space by emumerating tree topologies and/or discretising time.\\

To improve the ability to simulate the CwR, McVean and Cardin 
provided an approximation of the CwR that is both Markovian backwards in time 
as well as left-to-right along the genome \citep{mcvean_2005_approximating}. 
By sacrificing precision on our ability to capture long range linkage 
information, the sequentially Markovian coalescent (SMC) 
became the engine of many powerful inference methods [PSMC]. 
A later improvement, the SMC', closed the gap between the SMC and the 
CwR even further.

A second key approximation is not a generative model, instead the 
copying model \citep{li_2003_modeling} was specifically designed to 
generate CwR-like patterns. 
Rather than providing a mathematical description of the genealogies 
underlying the observed data it pragmatically describes the relationship 
of a single genome to a larger set of samples as the result of an 
incomplete copying process. This heuristic approach has been successfully 
applied for demographic inference [and other things: ADD] 
\citep{sheehan_2013_estimating, steinrucken_2018}.

Both approaches have further formed the basis of the recent advancements in 
scalable ARG inference methods [references]. 
First real breakthrough, ARGweaver, samples an ARG from the posterior distribution 
given the observed variational data. Although require limiting the set of 
possible node times to be compatible with HMM algorithms, the ARGs returned 
by ARGweaver are fully precise and detailed in the sense that all 
recombination events are explictly inferred. 
This is where the real big difference lies with all other scalable methods. 
Although all different, all these methods share the fact that the ARGs they infer 
are approximate structures. The ARGs they infer have varying degrees of detail

@@

% what is an approximate structure: implicit recombination, polytomies or not, ...
They lack the full detail: implicit recombination, ...
In particular, recombination is dealt with implicitly. Recombination events are 
not directly inferred and thus not dated. Tree topology changes from one local 
tree to the next without explicit inference of the one or more recombination 
events required to explain the inferred tree transition.

@@

% Existing formalisms assume that an ARG is known in full detail
These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).
